The Multi-Stencil Language

### 2016-05-20 ARTICLES/IJPP/article/msl.tex
IJPP article modifications plan + partie 2 + partie 3

\begin{filecontents*}{grammar.txt}
program ::= "mesh:" meshid 
            "mesh entities:" listmeshent
            "computation domains:" 
                      listcompdom
            "independent:"
                      listinde
            "stencil shapes:"
                      liststencil
            "data:" listdata
            "scalar" listscalar
            listloop

listmeshent ::= meshent listmeshent 
                        |  meshent
listcompdom ::= compdom listcompdom 
                        |  compdom
compdom ::= compdomid "in" listmeshent
listinde ::= inde listinde |  inde
inde ::= compdomid "and" compdomid
liststencil ::= stencil liststencil 
                        | stencil
stencil ::= stencilid "from" meshent "to" meshent
listdata ::= data listdata |  data
data ::= dataid "," meshent
listscalar ::= scalar listscalar 
                      | scalar

listloop ::= loop listloop | loop
loop ::=  "time:" iteration
          "computations:" listcomp

iteration ::= num | criteria
criteria ::= criteriaid "(" listscalarread ")"
listscalarread ::= scalar listscalarread 
                          |  scalar
listcomp ::= comp listcomp |  comp
comp ::= dataid "[" compdomid "]=" 
                    compid "({" 
                    listscalarread "},{" 
                    listdataread "})"
listdataread ::= dataread listdataread 
                          |  dataread
dataread ::= dataid "[" stencil "]" 
              |  dataid
\end{filecontents*}

\begin{figure}[!h]
  \hspace{5mm}
  \begin{minipage}[!h]{0.5\textwidth}
    {\lstinputlisting[basicstyle=\small,mathescape,frame=single,language=Python,numbers=left]{grammar.txt}}   
    \caption{Grammar of the Multi-Stencil Language. \label{fig:grammar}}
  \end{minipage}
\end{figure}

\begin{filecontents*}{exmsl.txt}
mesh : cart
mesh entities : cell, edgex
computation domains :
  d1 in cell
  d2 in edgex
independent :
  d1 and d2
stencil shapes : 
  n1 from cell to cell
  n2 from cell to edgex
  n3 from edgex to cell
data :
  A,cell
  B,cell
  C,edgex
  D,cell
  E,cell
  F,cell
  G,cell
  H,edgex
  I,cell
  J,cell
scalar : mu, tau
time : 500
computations :
  B[d1] = k0({tau},{A})
  C[d2] = k1({},{B[n2]})
  D[d1] = k2({},{C})
  E[d1] = k3({},{C})
  F[d1] = k4({},{D,C[n3]})
  G[d1] = k0({mu,tau},{E})
  H[d2] = k6({},{F})
  I[d1] = k7({},{G,H})
  J[d1] = k8({mu},{I[n1]})
\end{filecontents*}

\begin{figure}[!h]
  \hspace{5mm}
  \begin{minipage}[!h]{0.4\textwidth}
    {\lstinputlisting[basicstyle=\small,mathescape,frame=single,language=Python,numbers=right]{exmsl.txt}}   
    \caption{Example of program using the Multi-Stencil Language. \label{fig:exmsl}}
  \end{minipage}
\end{figure}

>>>>

From the fomalism detailed in the previous section, the Multi-Stencil Language and its grammar can already be given. The reason why this grammar is sufficient to automatically extract parallelism will be explained in the next section.

\medskip
The grammar of the Multi-Stencil Language is given in Figure~\ref{fig:grammar}. A Multi-Stencil program is composed of eight parts.

\begin{enumerate}
\item The first part represents the mesh of the simulation $\mathcal{M}$. As a single mesh is for now supported by the language, and as the topology of the mesh is not needed by the language compiler, \textit{meshid} (line 1) is a terminal symbol of the grammar. 
\item The second part (line 2) represents the set of groups of mesh entities needed by the simulation $\Phi$. It is defined as a list of groups $\mathcal{G}$. Again, as mesh entities are defined using the topology of the mesh, a group of mesh entities \textit{group} (defined line 13) is a terminal symbol of the grammar.
\item The third part (lines 3 and 4) represents the set of computation domains of the simulation $\mathcal{D}$. It is defined as a list of domains, each one defined (lines 14 and 15) as a subpart of a group of mesh entities, using the symbol "in".
\item The fourth part (lines 5 and 6) of the program is linked to the third one. It offers a way to declare that two computation domains are independants (lines 16 and 17). Two given computation domains $D_1$ and $D_2$ are independants if and only if $D_1 \cap D_2 = \emptyset$. It uses the symbol "and" to describe that $D_1$ and $D_2$ are independants.
\item The fith part (lines 7 and 8) represents the set of stencil shapes used by the simulation $\mathcal{N}$. A stencil shape (lines 18 and 19) is defined as a function "from" a group of mesh entities "to" another group of mesh entities.
\item The sixth part (line 9) represents the set of quantities of the simulation $\Delta$. It is composed of a list of quantities (lines 20 and 21) for which each line (line 22) begins with a group of mesh entities and a list of \textit{quanityid} which are applied onto the given group.
\item The seventh part (line 10) represents the set of scalars of the simulation $\mathcal{S}$. Each scalar (line 23) is a terminal symbol.
\item Finally, the eith part (line 11) represents the simulation computations. It is composed of a list of loop (line 24). Each \textit{loop} (line 25) is composed of a \textit{time} loop composed of \textit{iteration} (which represents $T$), and of a set of \textit{computations} (which represents $\Gamma$).
\begin{itemize}
\item A time loop, denoted \textit{iteration} in the grammar, can be a numerical constant value, \textit{num}, which directly indicates the number of time iterations, or a convergence \textit{criteria}. The \textit{criteria} symbol represents the function $conv:\mathcal{S}^n \rightarrow bool$ of the formalism. To define a criteria the terminal \textit{criteriaid} has to be set, and between bracket are indicated a list of scalars to read.
\item Each computation of the list of \textit{computations} follows the information of the formalism $k(S,R,(w,D),comp)$ except the $comp$ function which is an implementation concern. To be more easy to write the syntax is different though.
\end{itemize}
\end{enumerate}

\begin{filecontents*}{grammar.txt}
program ::= "mesh:" meshid 
            "mesh entities:" listgroup
            "computation domains:" 
                      listcompdom
            "independent:"
                      listinde
            "stencil shapes:"
                      liststencil
            "quantities:" listquantities
            "scalar" listscalar
            listloop

listgroup ::= group listgroup |  group
listcompdom ::= compdom listcompdom |  compdom
compdom ::= compdomid "in" group
listinde ::= inde listinde |  inde
inde ::= compdomid "and" compdomid
liststencil ::= stencil liststencil | stencil
stencil ::= stencilid "from" group "to" group
listquantities ::= data listquantities |  quantity
quantity ::= group listquantityid
listquantityid ::= quantityid listquantityid |  quantityid
listscalar ::= scalar listscalar | scalar
listloop ::= loop listloop | loop
loop ::=  "time:" iteration
          "computations:" listcomp
iteration ::= num | criteria
criteria ::= criteriaid "(" listscalarread ")"
listscalarread ::= scalar listscalarread |  scalar
listcomp ::= comp listcomp |  comp
comp ::= dataid "[" compdomid "]=" compid "({"listscalarread
                                          "},{"listdataread "})"
listdataread ::= dataread listdataread |  dataread
dataread ::= dataid "[" stencil "]" |  dataid
\end{filecontents*}

\begin{figure}[!h]
  \hspace{5mm}
  \begin{minipage}[!h]{0.98\textwidth}
    {\lstinputlisting[basicstyle=\small,mathescape,frame=single,language=Python,numbers=left]{grammar.txt}}   
    \caption{Grammar of the Multi-Stencil Language. \label{fig:grammar}}
  \end{minipage}
\end{figure}

\paragraph{\textbf{Example}} To give a better idea of a multi-stencil program, a short example is given in Figure~\ref{fig:exmsl}. In this example, the mesh is called \textit{cart}. Two groups of mesh entities are defined \textit{cell} and \textit{edgex}. As previously explained, it is not needed in the grammar to give details on the topology of the mesh and the group of mesh entities. Two computation domains are given, \textit{d1} is a subpart of the entities of the group \textit{cell}, and \textit{d2} is a subpart of the entities of the group \textit{edgex}. It is declared that \textit{d1} and \textit{d2} are independant. Three stencil shapes are given. The second one for example \textit{nce} returns mesh entities of the group \textit{edgex} from an entity of the group \textit{cell}. Eight quantities are applied onto the group \textit{cell}, and two onto the group \textit{edgex}. Two scalars are defined, \textit{mu} and \textit{tau}. The time loop will iterate 500 times. Finally, nine computations are declared. For example, the computation \textit{k8} write the quantity \textit{J} onto the computation domain \textit{d1} by reading the scalar \textit{mu} and the quantity \textit{I} with a stencil shape \textit{n1}.

\begin{filecontents*}{exmsl.txt}
mesh : cart
mesh entities : cell, edgex
computation domains :
  d1 in cell
  d2 in edgex
independent :
  d1 and d2
stencil shapes : 
  ncc from cell to cell
  nce from cell to edgex
  nec from edgex to cell
quantities :
  cell A,B,D,E,F,G,I,J
  edgex C,H
scalar : mu, tau
time : 500
computations :
  B[d1] = k0({tau},{A})
  C[d2] = k1({},{B[n2]})
  D[d1] = k2({},{C})
  E[d1] = k3({},{C})
  F[d1] = k4({},{D,C[n3]})
  G[d1] = k0({mu,tau},{E})
  H[d2] = k6({},{F})
  I[d1] = k7({},{G,H})
  J[d1] = k8({mu},{I[n1]})
\end{filecontents*}

\begin{figure}[!h]
  \hspace{5mm}
  \begin{minipage}[!h]{0.98\textwidth}
    {\lstinputlisting[basicstyle=\small,mathescape,frame=single,language=Python,numbers=left]{exmsl.txt}}   
    \caption{Example of program using the Multi-Stencil Language. \label{fig:exmsl}}
  \end{minipage}
\end{figure}

The aim of the Multi-Stencil Language is to offer a way to describe a numerical simulation based on explicit numerical schemes. The language is mesh-agnostic as no information is asked to the user about the actual topology of the mesh. Moreover, information that is usefull for the implementation is also split from the MSL description. For this reason, the language grammar is simple.

In the next section will be shown how parallelization can be extracted from this simple language and how an empty parallel skeleton of the application can be generated, introducing implementation concerns in a second phase.
